---
title: "An R package to access NOAA NCEI Blended Seawinds (NBS v2) "
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Provides for access to [NBS 2.0](https://coastwatch.noaa.gov/cwn/products/noaa-ncei-blended-seawinds-nbs-v2.html) from [NOAA Coastwatch](https://coastwatch.noaa.gov/cwn/index.html) which replaces the first version called [Blended Sea Winds](https://www.ncei.noaa.gov/products/blended-sea-winds) online.

### [Citation](https://doi.org/10.3389/fmars.2022.935549)

> Saha, K.; Huai-Min, Z. Hurricane and Typhoon Storm Wind Resolving NOAA NCEI Blended Sea Surface Wind (NBS) Product. Frontiers in Marine Sciences – Ocean Observation 2022, 9, 1–12. https://doi.org/10.3389/fmars.2022.935549.

### Requirements

From CRAN...

  + [R v4+](https://www.r-project.org/)
  + [rlang](https://CRAN.R-project.org/package=rlang)
  + [dplyr](https://CRAN.R-project.org/package=dplyr)
  + [tidyr](https://CRAN.R-project.org/package=tidyr)
  + [stars](https://CRAN.R-project.org/package=stars)
  + [sf](https://CRAN.R-project.org/package=sf)
  + [R6](https://CRAN.R-project.org/package=R6)
  

### Installation

```
remotes::install_github("BigelowLab/nbs")
```

```{r}
suppressPackageStartupMessages({
  library(nbs)
  library(dplyr)
  library(sf)
  library(stars)
})
```

#### Organization of Blended Seawinds

Blended Sea Winds are organized by product and interval.  The two primary product types are wind velocities and  wind stresses.  Each product type is organized further by aggregation interval `[6h, daily, monthly]`. 

```{r}
nbs_tally()
```


#### Working with points.  

See the [xyzt](https://github.com/BigelowLab/xyzt) package for more details on the example Southern US Atlantic Bight data.  Note we must transform the longitude values from [-180, 180] to [0,360] to work with BSW data.

```{r}
# read in example SAB points - note that time is required as a dimension
x <- xyzt::read_sab() |>
  dplyr::select(-time, -depth) |>
  dplyr::mutate(lon = xyzt::to_360(lon),
                time = as.POSIXct("1995-12-18 23:00:05", tz = "UTC")) |>
  xyzt::as_POINT(dims = "xyt")


# generate the BSW url for a given product
url <- bsw_url("Aggregation_of_6h_Ocean_Wind")


# open the resource (we can close it later)
X <- ncdf4::nc_open(url)

# extract the data 
covars <- bsw::extract(x, X, varname = c("u", "v"))

# bind to the input
(y <- dplyr::bind_cols(x, covars))
```

#### Working with bounding boxes (from points or polygons). 

Learn more about working with [stars](https://CRAN.R-project.org/package=stars) objects in the [vignettes](https://r-spatial.github.io/stars/).

```{r}
# read in example SAB points
x <- xyzt::read_sab() |>
  dplyr::select(-time, -depth) |>
  dplyr::mutate(lon = xyzt::to_360(lon),
                time = as.POSIXct("1995-12-18 23:00:05", tz = "UTC")) |> 
  xyzt::as_BBOX(dims = 'xyt')

(covars <- bsw::extract(x, X, varname = c("u", "v")))
```

Now let's see what it looks like.

```{r}
x <- xyzt::read_sab() |>
  dplyr::select(-time, -depth) |>
  dplyr::mutate(lon = xyzt::to_360(lon)) |>
  xyzt::as_POINT()
par(mfrow = c(1,2))
plot(covars, attr = 'u', axes = TRUE, reset = FALSE)
plot(sf::st_geometry(x), add = TRUE, col = "orange", pch = 19, cex = 2)
```



```{r}
# cleanup
ncdf4::nc_close(X)
```
