---
title: "An R package to access NOAA NCEI Blended Seawinds (NBS v2) "
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Provides for access to [NBS 2.0](https://coastwatch.noaa.gov/cwn/products/noaa-ncei-blended-seawinds-nbs-v2.html) from [NOAA Coastwatch](https://coastwatch.noaa.gov/cwn/index.html) which replaces the first version called [Blended Sea Winds](https://www.ncei.noaa.gov/products/blended-sea-winds) online.

This package provides tools for querying, downloading and archiving blended sea wind raster files. Data are served on THREDDS catalog using OPeNDAP services.  We hide all of the details and provide a simplified interface to download rasters (saved as GeoTIFF) and simple database organization.

### [Citation](https://doi.org/10.3389/fmars.2022.935549)

> Saha, K.; Huai-Min, Z. Hurricane and Typhoon Storm Wind Resolving NOAA NCEI Blended Sea Surface Wind (NBS) Product. Frontiers in Marine Sciences – Ocean Observation 2022, 9, 1–12. https://doi.org/10.3389/fmars.2022.935549.

## Requirements

From CRAN...

  + [R v4+](https://www.r-project.org/)
  + [rlang](https://CRAN.R-project.org/package=rlang)
  + [dplyr](https://CRAN.R-project.org/package=dplyr)
  + [tidyr](https://CRAN.R-project.org/package=tidyr)
  + [stars](https://CRAN.R-project.org/package=stars)
  + [sf](https://CRAN.R-project.org/package=sf)
  + [R6](https://CRAN.R-project.org/package=R6)
  + [thredds](https://CRAN.R-project.org/package=thredds)
  + [readr](https://CRAN.R-project.org/package=readr)
  + [rerddap](https://CRAN.R-project.org/package=rerddap)


## Installation

```
remotes::install_github("BigelowLab/nbs")
```

## Usage

```{r}
suppressPackageStartupMessages({
  library(nbs)
  library(dplyr)
  library(sf)
  library(stars)
})
```

### Set your root data path

Our workflow requires a path where you will save your requested data. We suggest that you store your root data path in a hidden configuration file (defaults to `~/.nbsdata`). This is a fix-and-forget step, but you can override or change it later. Here is how we set up my data path.

```
my_data_path = "/Users/ben/Library/CloudStorage/Dropbox/data/noaa/nbs"
ok = dir.create(my_data_path, showWarnings = FALSE, recursive = TRUE)
nbs::set_root_path(my_data_path)
```

That's it. When you later download data for some region, say for the Mediterranean, you can specify a subfolder, perhaps something like `med`, and store regional data there.

#### Organization of Blended Seawinds

Blended Sea Winds are organized by product and interval.  The two primary product types are wind velocities and  wind stresses.  Each product type is organized further by aggregation interval `[6h, daily, monthly]`. In addition, NOAA/NCEI serves up both "near real time" (NRT) and science-digest versions (SCI). These are served as "aggregate" or "per-file" OPeNDAP resources on the THREDDS server.  We find that accessing the "per-file" resources is more responsive than the otherwise more convenient "aggregate" resources.  

Here we tally the the 10 resources available to us. *Caveat* - we don't use the near real time (NRT) resources thus we haven't tested them.

```{r}
nbs_tally()
```
#### Query for a product

Use `nbs_query()` to search for product URLs.

```{r}
uris = query_nbs(period = "monthly",
                 dates = as.Date(c("1995-01-01", "1995-12-31")),
                 product = "uvcomp") |>
  dplyr::glimpse()
```
### Fetching data

When data are fetched they are automatically stored in GeoTIFF format in a database-like directory structure in the path of your own choosing. Each GeoTIFF file contains one parameter, at one depth and one time.  

We'll download just a portion of the western South Atlantic. First we need to make the destination path for the data, and then define the bounding box. Then we simply pass the table of URLs and the bounding box to `fetch_nbs()`.  We also provide the bounding box and path.  `param = "all"` is shorthand for fetching `c("u_wind", "v_wind", "windspeed")` from the `uvcomp` product.   

In return we'll receive a small table (*aka* database) listing the downloaded files.

```{r}
path = nbs_path("wsa")
ok = dir.create(path, recursive = TRUE, showWarnings = FALSE)
bb = c(xmin = 290, ymin = -60, xmax = 360, ymax = 10)
db = fetch_nbs(uris, params = 'all', bb = bb, path = path, verbose = FALSE)
```

### The database
The database contains identifying information about the stored files **except** for the path to the storage location.  This adds a little extra work for the end user (keeping track of the data path), but that extra effort is offset by improved portability and ease of use.

```{r}
db
```

We provide functions, `write_database()` and `read_database()`, to simplify the input-output.

```{r}
write_database(db, path)
```

### Reading in data

We can filter the database using standard tools, and then read in the data as rasters.

```{r}
db2 = dplyr::filter(db, param %in% c("u_wind", "v_wind"), 
                        date > as.Date("1995-10-01"),
                        date <= as.Date("1995-12-31"))
x = read_nbs(db2, path)
x
```
Note that the object has two attributes ("u_wind" and "v_wind") and three layers (October, November and December).

```{r plot_u}
plot(x['u_wind'])
```
```{r plot_v}
plot(x['v_wind'])
```