---
title: "An R package to access NOAA NCEI Blended Seawinds (NBS v2) "
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Provides for access to [NBS 2.0](https://coastwatch.noaa.gov/cwn/products/noaa-ncei-blended-seawinds-nbs-v2.html) from [NOAA Coastwatch](https://coastwatch.noaa.gov/cwn/index.html) which replaces the first version called [Blended Sea Winds](https://www.ncei.noaa.gov/products/blended-sea-winds) online.

### [Citation](https://doi.org/10.3389/fmars.2022.935549)

> Saha, K.; Huai-Min, Z. Hurricane and Typhoon Storm Wind Resolving NOAA NCEI Blended Sea Surface Wind (NBS) Product. Frontiers in Marine Sciences – Ocean Observation 2022, 9, 1–12. https://doi.org/10.3389/fmars.2022.935549.

### Requirements

From CRAN...

  + [R v4+](https://www.r-project.org/)
  + [rlang](https://CRAN.R-project.org/package=rlang)
  + [dplyr](https://CRAN.R-project.org/package=dplyr)
  + [tidyr](https://CRAN.R-project.org/package=tidyr)
  + [stars](https://CRAN.R-project.org/package=stars)
  + [sf](https://CRAN.R-project.org/package=sf)
  + [R6](https://CRAN.R-project.org/package=R6)
  

### Installation

```
remotes::install_github("BigelowLab/nbs")
```

```{r}
suppressPackageStartupMessages({
  library(nbs)
  library(dplyr)
  library(sf)
  library(stars)
})
```

#### Organization of Blended Seawinds

Blended Sea Winds are organized by product and interval.  The two primary product types are wind velocities and  wind stresses.  Each product type is organized further by aggregation interval `[6h, daily, monthly]`. In addition, NOAA/NCEI serves up both "near real time" (NRT) and science-digest versions (SCI).

```{r}
nbs_tally()
```


#### Opening a session with a product

```{r}
X = NBS2$new(product = 'uvcomp_SCIMonthlyGlobal')

times = X$get_time()
head(times)
tail(times)
```

#### Extracting with bounding boxes

The grid is a regular WGS84 `longlat` grid, but you should note that the longitudes range from (0, 360) and not (-180, 180).

You can either manually specify a bound box in (xmin, ymin, xmax, ymax) order, or provide an `sf-like` object from which the bounding box is extracted. Below we extract 12 months (from 2020), bind as one stars object and display.

```{r}
bb = c(290, -60, 360, 10)
dates = seq(from = as.Date("2020-01-01"), 
            to = as.Date("2020-12-01"), 
            by = "month")
windspeeds = lapply(seq_along(dates),
                function(imonth){
                  X$get_var(bb = bb, varid = "windspeed", 
                            form = 'stars', time = dates[imonth])
                }) 
windspeed = do.call(c, append(windspeeds, list(along = list(date = dates))))
plot(windspeed)
```

#### The pesky [-180, 180] vs [-0, 360] issues

It seems like most mapping paradigms we see are mapped in the range [-180, 180], but certainly not all.  If you have a bounding box using the [-180, 180] style, then simply convert it using the provided functionality.  Below we show how to transform a bounding box, make the extraction and then 'shift' the result to [-180, 180] space.

```{r}
bb_180 = c(-74, 39, -55, 50)
bb_360 = bb_to360(bb_180)
windspeed_360 = X$get_var(bb = bb_360, time = as.Date("1995-12-01"), varid = "windspeed")
sf::st_bbox(windspeed_360)
plot(windspeed_360, axes = TRUE)
```
Now we can shift the stars object to be in the [-180, 180] space.
```{r}
windspeed_180 = stars_to180(windspeed_360)
sf::st_bbox(windspeed_180)
plot(windspeed_180, axes = TRUE)
```

#### Crossing the edge

How does one get a single object if the bounding box crosses the edge?  It's simple, really, to split the bounding box into two, extract each separately and then use `stars::st_mosaic()` to merge them into a single object.

First we'll make a bb in [-180, 180] space, then split it into 2 at 0 degrees.
```{r}
bb_over_edge_180 = c(-30, 10, 30, 50)
bbs_180 = bb_split(bb_over_edge_180, at = 0)
bbs_180
```
We convert each to [0,360].

```{r}
bbs_360 = lapply(bbs_180, bb_to360)
bbs_360
```

Now we collect a [0,360] stars object for each bb, shifting to [-180, 180] as needed before mosaicing.

```{r}
xx = lapply(bbs_180,
            function(bb_180){
              X$get_var(time = as.Date("1989-01-01"),
                        varid = "windspeed",
                        bb= bb_to360(bb_180)) |>
                stars_to180()
            })
windspeed = stars::st_mosaic(xx[[1]], xx[[2]]) 
plot(windspeed, axes = TRUE)
```
            }


